An algorithm for a spatial mapping using a hexagon tilegram, with application to Australia disease maps

### 2019-01-07 algorithm_radix.Rmd
Parameters section

- Like Dorling, same shape will be used.
Like tilegram, tesselation using equally sized shapes.

- Allocate each geographic area to a map space of consistent tesselated shape

which is different from existing algorithms:
- contiguous, keeps exact neightbours, no recognisable shape
- non contiguous, keeps shape, can lose neighbours

@@

@@

To utilise a shape file from a directory on your device, provide the file path for .shp file.
The `st_read` function is applied internally to create the sf object.
If the shape file is an .Rda file, it is loaded into the current environment.
Empty geometries are removed from the sf object, and the remaining polygons are projected to the Australian standard.

```{r}

```

The geometry column now contains projected polygons,
the projected centroids are derived for all polygons remaining in the set.

## Hexgon grid
Create a grid of hexagon centroids using the `create_grid` function.

The following parameter values are necessary for this step, and will be automatically derived in the `create_hexmap` function if not provided:

- buffer distance: *the distance, in degrees, to extend the hexagon grid beyond the polygon centroids provided*
- size of the hexagons: *a float value in degrees for the width of the hexagons to be placed on the map space*
- filter distance: *the amount of hexagons around each centroid to consider for allocation, by default this is 10 hexagons*

The grid creation takes several steps:

```{r}
# create grid calls these functions
# call create grid here, or call these parts individually, may be a lot of code
```

@@

This grid is square, which will not facilitate tesselated hexagons. 
Every second latitude (row) of points will be shifted right by half of the hexagon size.

@@

### Step 2: Rolling windows
The longitude columns and latitude rows are given integer id values.
The rows and columns are then divided into 20 groups, 
recording the amount of rows in each latitude group and the amount of columns in each longitude group.

The amount of rows and columns are then used to make manual rolling windows.

The first rolling window function finds the minimum and maximum centroid values for the longitude columns and latitude rows.

@@

### Step 3: Filtering the grid
Only the grid points between the rolling average of the minimum and maximum centroid valuesare kept, for each row and column of the grid.

@@

If no available hexagon grid point is found within the original filter distance and angle, the distance is expanded, only when a maximum distance is reached will the angle expand to accomodate more possible grid points.

@@

<!--
Prosecode:  Number each step, never break a loop over steps,
use subnumbering for parts of each step, and
include explanatory text.
Literate code:  The algorithm is introduced gradually,
intermingled with discussion of underlying ideas.
-->

@@

Examples of Melbourne, Victoria, Australia
At different levels, distance from original centroids, change in area

LGA of all australia
sa2 for queensland

@@

Areal Statistic Distortion is the change in the areal distribution of the set of polygons.
Difference between space taken by polygons and space taken by hexagons, with respect to the desired statistic. Setting the desired statistic to area givesthe change in map space used to represent the polygons.

@@

<!--
Apply methods of evaluation used elsewhere
Nusrat, Alam and Kobourov:
- Statistical accuracy: how well do the modified areas
represent the corresponding statistic shown (e.g.,
population or GDP). This is measured in terms of
“cartographic error.”
- Geographical accuracy: how much do the modified
shapes resemble the original geographic shapes and
how well preserved are their relative positions.
- Topological accuracy: how well does the topology (as
measured by adjacent regions) of the cartogram
match that of the original map.
-->

>>>>

- Like Dorling, same shape will be used.
Like tilegram, tessellation using equally sized shapes.

- Allocate each geographic area to a map space of consistent tessellated shape

which is different from existing algorithms:
- contiguous, keeps exact neighbours, no recognisable shape
- non contiguous, keeps shape, can lose neighbours

@@

## Parameters

The `create_hexmap` functions requires several parameters. If not provided the information may be derived from the polygon set used.

**The following must be provided to `create_hexmap`:**

The set of polygons to be made into a hexagon map:
`shp: a shape file`

The location of a set of polygons to be made into a hexagon map:
`shp_path: character string location of shape file`

A column of unique names or id values for each area
`sf_id: name of a unique column that distinguishes areas`

A data frame of reference points to determine the order of allocation
`focal_points a data frame of reference locations when allocating hexagons, capital cities of Australia are provided as capital_cities`

```{r}
library(tidyverse)
library(sugaRbag)

shp_path <- "data/lga_2011.rda"
sf_id <- "LGA_CODE11"
focal_points <- sugaRbag::capital_cities
```

**The following will be determined within `create_hexmap` if not provided, but must be created for the following example:**

- `projstring	a string to indicate the projection and epsg`
- `epsg	the four character string to indicate the CRS`
- `buffer_dist: a float value for distance in degrees to extend beyond the - geometry provided`
- `hex_size	a float value in degrees for the diameter of the hexagons`
- `filter_dist amount of hexagons around centroid to consider for allocation`
- `width the angle used to filter the grid points around a centroid`

These parameters will be created when they are needed in the example:

```{r}
width <- 30 # arbitrary angle width
```

<!--
needed later:

- filter distance: *the amount of hexagons around each centroid to consider for allocation, by default this is 10 hexagons*
-->

@@

- To utilise a shape file from a directory on your device, provide the file path for .shp file.
- The `st_read` function is applied internally to create the sf object.
- If the shape file is an .Rda file, it is loaded into the current environment.

Empty geometries are removed from the sf object, and the remaining polygons are projected to the Australian standard.

We will use the Local Government Areas of Hobart as an example:

```{r}
# Read in shape information
tas_shp <- read_shape(shp_path = shp_path)

# Convert to a simple features object
tas_lga <- sf::st_as_sf(tas_shp)

# Select only desired LGA areas, those from Tasmania.
tas_lga <- tas_shp %>% filter(STE_NAME11 == "Tasmania")
tas_lga[[sf_id]] <- droplevels(as.factor(tas_lga[[sf_id]]))

## Not necessary within `create_hexmap`:
## Derive projection information
crs_info <- sf::st_crs(tas_lga)
proj4string <- crs_info$proj4string
epsg <- crs_info$epsg

## For UTM as per shape file
projstring <- paste0("+init=epsg:", epsg, " ", proj4string, collapse = "")
## if long lat is preferred to UTM, and not in proj4string:
projstring <- "+init=epsg:3112 +proj=longlat +lat_1=-18 +lat_2=-36 +lat_0=0 +lon_0=134 +x_0=0 +y_0=0 +ellps=GRS80 +towgs84=0,0,0,0,0,0,0 +no_defs"

## Use the create_centroids function to find polygon centroids:
centroids <- create_centroids(shp_sf = tas_lga, sf_id = sf_id, projstring = projstring, verbose = FALSE)
```

## Hexagon grid
Create a grid of hexagon centroids using the `create_grid` function.

The grid creation takes several steps. It requires the centroids, the bounding box, the hexagon size and the buffer distance.

```{r}
## Not necessary within `create_hexmap`:
## create the bounding box from the set of polygon centroids:
bbox <- tibble::tibble(min = c(min(centroids$longitude), min(centroids$latitude)),
        max = c(max(centroids$longitude), max(centroids$latitude)))

hex_size <- (bbox$max[1] - bbox$min[1])/(bbox$max[2] - bbox$min[2]) / 10

buffer_dist <- max((bbox$max[1] - bbox$min[1]), (bbox$max[2] - bbox$min[2]))*0.3
```

@@

```{r}
grid <- tibble::as.tibble(expand.grid(hex_long = seq(bbox$min[1] - buffer_dist,
    bbox$max[1] + buffer_dist,
    hex_size),
    hex_lat = seq(bbox$min[2] - buffer_dist,
        bbox$max[2] + buffer_dist,
        hex_size)))
```

This grid is square, which will not facilitate tessellated hexagons. 
Every second latitude (row) of points will be shifted right by half of the hexagon size.

```{r}
# Find every second latitude
shift_lat <- grid %>% dplyr::select(hex_lat) %>%
    dplyr::distinct() %>%
    dplyr::filter(dplyr::row_number() %% 2 == 1) %>% unlist()

# Shift the longitude of every second latitude to the right to make hex structure
grid <- grid %>%
    dplyr::mutate(hex_long = ifelse(hex_lat %in% shift_lat, hex_long,
        hex_long + (hex_size / 2))) %>%
    dplyr::mutate(id=1:NROW(.)) %>%
    dplyr::mutate(assigned=FALSE)
```

Each row and column is given an ID

```{r}
grid <- grid %>%
    mutate(hex_long_int = dense_rank(hex_long)-1,
        hex_lat_int = dense_rank(hex_lat)-1)
```

@@

```{r}
return_grid <- create_buffer(centroids, grid, hex_size, buffer_dist)
```

### Step 2: Rolling windows
The rows and columns are divided into 20 groups. The amount of rows in each latitude group and the amount of columns in each longitude group are then used as the width of manual rolling windows.

The first rolling window function finds the minimum and maximum centroid values for the groups of longitude columns and the groups of latitude rows.

@@

### Step 3: Filtering the grid
Only the grid points between the rolling average of the minimum and maximum centroid values are kept, for each row and column of the grid.

@@

If no available hexagon grid point is found within the original filter distance and angle, the distance is expanded, only when a maximum distance is reached will the angle expand to accommodate more possible grid points.

@@

<!--
Prose code:  Number each step, never break a loop over steps,
use sub numbering for parts of each step, and
include explanatory text.
Literate code:  The algorithm is introduced gradually,
intermingled with discussion of underlying ideas.
-->

@@

Examples of Melbourne, Tasmania, Australia
At different levels, distance from original centroids, change in area

LGA of all Australia
SA2 for Queensland

@@

Areal Statistic Distortion is the change in the areal distribution of the set of polygons.
Difference between space taken by polygons and space taken by hexagons, with respect to the desired statistic. Setting the desired statistic to area gives the change in map space used to represent the polygons.

@@

Old neighbours over new neighbours?
<!--
Apply methods of evaluation used elsewhere
Nusrat, Alam and Kobourov:
- Statistical accuracy: how well do the modified areas
represent the corresponding statistic shown (e.g.,
population or GDP). This is measured in terms of
“cartographic error.”
- Geographical accuracy: how much do the modified
shapes resemble the original geographic shapes and
how well preserved are their relative positions.
- Topological accuracy: how well does the topology (as
measured by adjacent regions) of the cartogram
match that of the original map.
-->
